---
title: "Macroalgal and mobile macro-invertebrate species face high extinction risk - part 3: plotting"
author: "Olivia J. Johnson, Freddie J. Heather & Camille Mellin"
date: "`r Sys.Date()`"
output: 
  rmdformats::downcute:
  # bookdown::html_document2:
    code_folding: show
---

# About this script

The following document provides the R code for the analysis for the paper: Johnson *et al.* (in review.) Macroalgal and mobile macro-invertebrate species face high extinction risk

**\*Corresponding author:**
[olivia.johsnon\@utas.edu.au](mailto:olivia.johsnon@utas.edu.au){.email}

+------------------------+-------------------------------------------------------------------------------------------------------+
| **R-version**          | 4.4.1 (2024-06-14 ucrt) -- "Race for Your Life"                                                        |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **platform**           | x86_64-w64-mingw32/x64                                                                                   |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article DOI**        |                                                                                                       |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article link**       |                                                                                                       |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article citation**   | Johnson *et al.* (in review.) Macroalgal and mobile macro-invertebrate species face high extinction risk |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Time series**        | 1992 - 2024                                                                                              |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Geographical scale** | Continental Australia                                                                                 |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Code author contact**| olivia.johnson@utas.edu.au                                                                            |
+------------------------+-------------------------------------------------------------------------------------------------------+

# Set-up

```{r knitr-options}
#| include = FALSE

library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# save csv files for each step?
save_csv <- TRUE

```

# Loading packages
Load in all required packages for this analysis
```{r packages}

library(knitr)
library(tidyverse)
library(janitor)
library(lubridate)
library(lme4)
library(lmerTest)
library(dplyr)
library(data.table)
library(zoo, include.only = "na.approx")
library(ggforce, include.only = "facet_col")
library(data.table)
library(vcd)

# "not in" function
`%!in%` <- Negate(`%in%`)
```

#Read in summarised results spreadsheet
```{r - read in data}
alltime_change <- read_csv("LMM_output.csv")
```

#Plotting
Read in csv of LMM results, rate of change from above and the species current 
IUCN status.This spreadsheet has also filtered out all of the species that 
produced the 'isSingular' error in the modelling stage. 
```{r}
alltime_change <- alltime_change %>%
  mutate(biogeog = recode(biogeog, 
                          `cool (temperate)` = "temperate (cool)", 
                          `warm (temperate)` = "temperate (warm)")) 

#Decade decline plots
alltime_change$taxon <- factor(alltime_change$taxon, levels = c("Macroalgae", "Invertebrate", "Vertebrate"))
alltime_change$biogeog <- factor(alltime_change$biogeog, levels = c("temperate (cool)", "temperate (warm)", "tropical"))


#Endemic species plot
# Define the colors used for the `geom_rect` layers
fill_colors <- c("#cccc33", "yellow", "orange", "red")

endemicdecade <- alltime_change %>%
  # Combine taxon and biogeog factor conversion into a single mutate call
  mutate(
    taxon = case_when(
      taxon == "Vertebrate" ~ "c) Fishes",
      taxon == "Invertebrate" ~ "b) Invertebrates",
      taxon == "Macroalgae" ~ "a) Macroalgae"
    ),
    taxon = factor(taxon, levels = c("a) Macroalgae", "b) Invertebrates", "c) Fishes")),
    biogeog = factor(biogeog, levels = c("tropical", "temperate (warm)", "temperate (cool)")),
    sig_filled = factor(if_else(str_detect(sig, "\\*"), "Significant", "Non-significant"), 
                        levels = c("Non-significant", "Significant")) # Ensure factor levels
  ) %>%
  # Include species with a decline of >= -25% in decade_change
  filter(decade_change <= -24.9, slope < 0, endemism == "endemic") %>%
  ggplot(aes(x = decade_change, y = reorder(species_name, -decade_change))) +
  # Define the background rectangles with xmin and xmax values
  geom_rect(ymin = -Inf, ymax = Inf, xmin = 0, xmax = -30, fill = "#cccc33") +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -30, xmax = -50, fill = "yellow") +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -50, xmax = -80, fill = "orange") +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -80, xmax = -100, fill = "red") +
  # Add vertical dashed lines
  geom_vline(xintercept = c(-30, -50, -80), lty = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  # Error bars with the same position nudge as the points
  geom_errorbarh(aes(xmin = decade_change_lwr, xmax = pmin(decade_change_upr, 0), height = 0.2),
                 position = position_nudge(y = -0.12)) +
  # Points with the same position nudge
  geom_point(aes(shape = biogeog, fill = sig_filled), size = 4, color = "black", stroke = 1, 
             position = position_nudge(y = -0.12)) +
  # Add facets for taxa with free y scale and spacing
  ggforce::facet_col(taxon ~ ., scales = "free_y", space = "free") +
  # Customize shapes for the different biogeographies
  scale_shape_manual(values = c(
    "temperate (cool)" = 16,
    "temperate (warm)" = 17,
    "tropical" = 15
  )) +
  # Customize fill for significance with drop = FALSE
  scale_fill_manual(
    values = c("Non-significant" = "transparent", "Significant" = "black"), 
    labels = c("Non-significant", "Significant"),
    drop = FALSE, # Prevents dropping missing levels
    guide = guide_legend(override.aes = list(fill = c("transparent", "black"), shape = 21))
  ) +
  # Axis labels and legend title
  labs(
    x = "Rate of decline over a decade",
    y = "Species Name",
    shape = "Biogeography",
    fill = "Significance"
  ) +
  # Apply minimal theme and customizations
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major.y = element_blank(),  # Remove grid lines for y-axis
    axis.title.y = element_blank(),        # Remove y-axis title
    axis.text.y = element_text(face = "italic"),  # Italicize species names
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 11, hjust = 0, vjust = 1)  # Facet label style
  )


plot(endemicdecade)

#Save plot of endemics
pdf("alltime_change_EndemicsDecade_Final.pdf", width = 8, height = 12)  # Dimensions in inches
print(endemicdecade)  # Replace 'map' with your plot object
dev.off()  # Close the device

#####################


##########Non-endemic species plot

# Define the colors used for the `geom_rect` layers
fill_colors <- c("#cccc33", "yellow", "orange", "red")

nonendemicsdecade <- alltime_change %>%
  # Combine taxon and biogeog factor conversion into a single mutate call
  mutate(
    taxon = case_when(
      taxon == "Vertebrate" ~ "c) Fishes",
      taxon == "Invertebrate" ~ "b) Invertebrates",
      taxon == "Macroalgae" ~ "a) Macroalgae"
    ),
    taxon = factor(taxon, levels = c("a) Macroalgae", "b) Invertebrates", "c) Fishes")),
    biogeog = factor(biogeog, levels = c("tropical", "temperate (warm)", "temperate (cool)")),
    # Define a binary for significance (filled vs hollow shapes)
    sig_filled = if_else(str_detect(sig, "\\*"), "Significant", "Non-significant")
  ) %>%
  # Include species with a decline of >= -25% in decade_change
  filter(decade_change <= -24.9, slope < 0, endemism == "non-endemic") %>%
  ggplot(aes(x = decade_change, y = reorder(species_name, -decade_change))) +
  # Define the background rectangles with xmin and xmax values
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -0, xmax = -30, fill = fill_colors[1]) +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -30, xmax = -50, fill = fill_colors[2]) +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -50, xmax = -80, fill = fill_colors[3]) +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -80, xmax = -100, fill = fill_colors[4]) +
  
  # Add vertical dashed lines
  geom_vline(xintercept = c(-30, -50, -80), lty = 2) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  
  # Error bars with the same position nudge as the points
  geom_errorbarh(aes(xmin = decade_change_lwr, xmax = pmin(decade_change_upr, 0), height = 0.2),
                 position = position_nudge(y = -0.12)) +
  
  # Points with the same position nudge
  geom_point(aes(shape = biogeog, fill = sig_filled), size = 4, color = "black", stroke = 1, 
             position = position_nudge(y = -0.12)) +
  
  # Add facets for taxa with free y scale and spacing
  ggforce::facet_col(taxon ~ ., scales = "free_y", space = "free") +
  
  # Customize shapes for the different biogeographies
  scale_shape_manual(values = c(
    "temperate (cool)" = 16,
    "temperate (warm)" = 17,
    "tropical" = 15
  )) +
  
  # Customize fill for significance (clear for non-significant, black for significant)
  scale_fill_manual(values = c("Significant" = "black", "Non-significant" = "transparent"), 
                    labels = c("Non-significant", "Significant"), 
                    guide = guide_legend(override.aes = list(fill = c("transparent", "black"), shape = 21))) +
  
  # Axis labels and legend title
  labs(
    x = "Rate of decline over a decade",
    y = "Species Name",
    shape = "Biogeography",
    fill = "Significance"
  ) +
  
  # Apply minimal theme and customizations
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid.major.y = element_blank(),  # Remove grid lines for y-axis
    axis.title.y = element_blank(),        # Remove y-axis title
    axis.text.y = element_text(face = "italic"),  # Italicize species names
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 11, hjust = 0, vjust = 1)  # Facet label style
  )


plot(nonendemicsdecade)

#Save plot of non-endemics
pdf("alltime_change_NonendemicsDecade_Final.pdf", width = 8, height = 12)  # Dimensions in inches
print(nonendemicsdecade)  # Replace 'map' with your plot object
dev.off()  # Close the device

```

#IUCN Status versus trend plots
```{r}
#Specify order of facet panels in plots
#desired_order <- c("Macroalgae", "Coral", "Invertebrate", "Vertebrate")

#ENDEMICS
all_species_endemic <- alltime_change %>%
    filter(endemism == "endemic") %>%
 mutate(provisional_status = case_when(
    decade_change >= -14.99 ~ "Least Concern",
    decade_change >= -29.99 & decade_change < 15 ~ "Near Threatened",
    decade_change >= -49.99 & decade_change < -30 ~ "Vulnerable",
    decade_change >= -79.99 & decade_change < -50 ~ "Endangered",
    decade_change >= -100 & decade_change < -80 ~ "Critically Endangered"
  )) 

all_species_endemic <- all_species_endemic  %>%
  count(taxon, IUCN, provisional_status, species_name) %>%
  mutate(provisional_status = factor(provisional_status, levels = c("Critically Endangered", "Endangered", "Vulnerable", "Near Threatened", "Least Concern", "Lower Risk/conservation dependent"))) %>%
  mutate(taxon = case_when(taxon == "Vertebrate" ~ "c) Fishes",
                           taxon == "Invertebrate" ~ "b) Invertebrates",
                           taxon == "Macroalgae" ~ "a) Macroalgae")) %>%
  mutate(taxon = factor(taxon, levels = c("a) Macroalgae", "b) Invertebrates", "c) Fishes"))) %>%
  ggplot(aes(x = IUCN, y = n, fill = provisional_status)) +
  geom_col() +
  facet_wrap(~taxon, ncol = 1,scales = "free_y") + 
  labs(x = "IUCN Red List Conservation Status",
       y = "Species Count",
       fill = "Population trend") +
  scale_fill_manual(
        values = c("Critically Endangered" = "red", "Endangered" = "orange", "Vulnerable" = "yellow", "Near Threatened" = "#cccc33", "Least Concern" = "green3"),
    breaks = c("Critically Endangered", "Endangered", "Vulnerable", "Near Threatened", "Least Concern")) +
  # scale_x_discrete(limits = c("Not Evaluated", "Data Deficient", "Near Threatened", "Least Concern", "Vulnerable", "Endangered")) +
  scale_x_discrete(limits = c("Not Evaluated", "Data Deficient", "Least Concern","Near Threatened",  "Vulnerable", "Endangered")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    legend.position = "right",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 11, hjust = 0),
    axis.title.y = element_text(margin = margin(r = 20)),
    strip.text.x = element_text(margin = margin(b = 10)),
    strip.text.y = element_text(margin = margin(r = 10)),
    axis.title.x = element_text(margin = margin(r = 40, t = 20))
  )


#Save plot of IUCN endemics
pdf("alltime_change_EndemicsIUCN_Final.pdf", width = 6, height = 8)  # Dimensions in inches
print(all_species_endemic)  # Replace 'map' with your plot object
dev.off()  # Close the device

################################ NON-ENDEMICS #######################################

#NON-ENDEMICS

all_species_non_endemic <- alltime_change %>%
    filter(endemism == "non-endemic") %>% 
  mutate(provisional_status = case_when(
    decade_change >= -14.99 ~ "Least Concern",
    decade_change >= -29.99 & decade_change < 15 ~ "Near Threatened",
    decade_change >= -49.99 & decade_change < -30 ~ "Vulnerable",
    decade_change >= -79.99 & decade_change < -50 ~ "Endangered",
    decade_change >= -100 & decade_change < -80 ~ "Critically Endangered"
  )) 

#Specify order of facet panels in plots
#desired_order <- c("Macroalgae", "Coral", "Invertebrate", "Vertebrate")

all_species_non_endemic <- all_species_non_endemic %>%
  count(taxon, IUCN, provisional_status, species_name) %>%
  mutate(provisional_status = factor(provisional_status, levels = c("Critically Endangered", "Endangered", "Vulnerable", "Near Threatened", "Least Concern", "Lower Risk/conservation dependent"))) %>%
  mutate(taxon = case_when(taxon == "Vertebrate" ~ "c) Fishes",
                           taxon == "Invertebrate" ~ "b) Invertebrates",
                           taxon == "Macroalgae" ~ "a) Macroalgae")) %>%
  mutate(taxon = factor(taxon, levels = c("a) Macroalgae", "b) Invertebrates", "c) Fishes"))) %>%
  ggplot(aes(x = IUCN, y = n, fill = provisional_status)) +
  geom_col() +
  facet_wrap(~taxon, ncol = 1,scales = "free_y") + 
  labs(x = "IUCN Red List Conservation Status",
       y = "Species Count",
       fill = "Population trend") +
  scale_fill_manual(
    values = c("Critically Endangered" = "red", "Endangered" = "orange", "Vulnerable" = "yellow", "Near Threatened" = "#cccc33", "Least Concern" = "green3"),
    breaks = c("Critically Endangered", "Endangered", "Vulnerable", "Near Threatened", "Least Concern")) +
  # scale_x_discrete(limits = c("Not Evaluated", "Lower Risk/conservation dependent", "Data Deficient", "Least Concern", "Near Threatened", "Vulnerable", "Endangered")) +
  scale_x_discrete(limits = c("Not Evaluated", "Least Concern", "Near Threatened", "Vulnerable", "Endangered")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    legend.position = "right",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 11, hjust = 0),
    axis.title.y = element_text(margin = margin(r = 20)),
    strip.text.x = element_text(margin = margin(b = 10)),
    strip.text.y = element_text(margin = margin(r = 10)),
    axis.title.x = element_text(margin = margin(r = 40, t = 20))
  )

#Save plot of IUCN endemics
pdf("alltime_change_NonendemicsIUCN_Final.pdf", width = 6, height = 8)  # Dimensions in inches
print(all_species_non_endemic)  # Replace 'map' with your plot object
dev.off()  # Close the device
```


#Date of assessment
```{r - bar plot of evaluation dates}

yearassessed <- alltime_change %>%
  select(species_name, IUCN, date_assessed) %>%
  mutate(year_assessed = case_when(
    date_assessed == "Not Evaluated" ~ "Not Evaluated",  # Assign "Not Evaluated" if date_assessed is "Not Evaluated"
    !is.na(date_assessed) & date_assessed != "" ~ {
      parsed_date <- parse_date_time(date_assessed, orders = c("dmy", "mdy"), quiet = TRUE)
      as.character(year(parsed_date))  # Extract the year and convert to character
    },
    TRUE ~ "Not Evaluated"  # Assign "Not Evaluated" for all other NA cases
  ))

#Summarise the number of IUCN assessments in a given year
yearassessed <- yearassessed %>%
  group_by(year_assessed) %>%             
  summarise(total_species = n(),           
            .groups = 'drop') 


# Create the bar plot with minimal theme and black fill
yearassessed_plot <- ggplot(yearassessed, aes(x = factor(year_assessed, levels = unique(year_assessed)), y = total_species)) +
  geom_bar(stat = "identity", fill = "black", color = "black", width = 0.8) +
  labs(x = "Assessment Year",
       y = "Number of Assessments") +
  theme_minimal(base_size = 15) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    axis.title = element_text(face = "bold", size = 12),  # Adjust size here
    axis.text = element_text(color = "black"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_blank(),
    plot.background = element_blank(),
    axis.line = element_line(color = "black") 
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_x_discrete(expand = c(0, 0))

pdf("Year_assessed.pdf", width = 9, height = 10)  # Dimensions in inches
print(yearassessed_plot)  # Replace 'map' with your plot object
dev.off()  # Close the device


#There are 342 species "Not Evaluated" - 47% of common shallow reef species not assessed on the IUCN
```


#Produce map of sites where all data is available from around Australia
#MAP
```{r packages for mapping}
library(dplyr)  
library(ggplot2)
library(ggspatial)
library(ozmaps)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(rnaturalearthhires)
library(ggspatial)
library(tidyr)
```

```{r - read in data}
#Algae
algae_mappingsites <- read_csv("algae_mappingsites_2024.csv")

#Fish
fish_mappingsites <- read_csv("fish_mappingsites_2024.csv")

#Inverts
inverts_mappingsites <- read_csv("inverts_mappingsites_2024.csv")

#Coral
coral_mappingsites <- read_csv("coral_mappingsites_2024.csv")

```

```{r - themes}
theme_set(theme_bw())
world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)
```

#Determine latitude and longitude points
```{r - determine lats and longs - Clean}
#algae
algae_lats_longs_clean <- algae_mappingsites  %>%
  select(latitude, longitude, site_code, species_name)

#fish
fish_lats_longs_clean <- fish_mappingsites %>%
  select(latitude, longitude, site_code, species_name)

#inverts
inverts_lats_longs_clean <- inverts_mappingsites %>%
  select(latitude, longitude, site_code, species_name)

#coral
coral_lats_longs_clean <- coral_mappingsites %>%
  select(latitude, longitude, site_code, species_name)

# Combine latitude and longitude tables for algae, fish, inverts, and coral
all_lats_longs_clean <- bind_rows(algae_lats_longs_clean, fish_lats_longs_clean, inverts_lats_longs_clean, coral_lats_longs_clean) %>%
  select(latitude, longitude)

#Remove duplicates
all_lats_longs_clean <- all_lats_longs_clean %>%
  select(latitude, longitude) %>%
  distinct()

all_lats_longs_clean <- all_lats_longs_clean[!duplicated(all_lats_longs_clean[, c("latitude", "longitude")]), ]
```

#Produce map
```{r - lats and long code for map}
#RAW data map
coordinates <- all_lats_longs_clean #(above)

#coordinates (869 sites) used in analysis
coords_from_analysis_sites <- read_csv("filtered_site_coords_nov.csv")

#how many sites in total are on this map?
unique_species_count <- length(unique(coordinates$latitude))
unique_species_count #2652 sites

# Load world and Australian states data
world <- ne_countries(scale = "medium", returnclass = "sf")
oz_states <- ne_states(country = "Australia", returnclass = "sf")

#read in covariates
covariates <- read_csv("covariates_final.csv")

# Select relevant columns from covariates
covariates_selected <- covariates %>%
  select(Latitude, Longitude, temperature_mean_2020) %>%
  rename(latitude= Latitude, longitude = Longitude)

# Merge coordinates with temperature data
merged_data <- coordinates %>%
  left_join(covariates_selected, by = c("latitude", "longitude"))

# Assign temperature categories
merged_data <- merged_data %>%
  mutate(temperature_category = case_when(
    temperature_mean_2020 < 17.5 ~ "temperate (cool)",
    temperature_mean_2020 >= 17.5 & temperature_mean_2020 < 23 ~ "temperate (warm)",
    temperature_mean_2020 >= 23 ~ "tropical",
    TRUE ~ NA_character_  # Set to NA instead of "Unknown"
  ))

# Fill down the temperature categories for NAs
merged_data <- merged_data %>%
  arrange(latitude, longitude) %>%  # Ensure data is sorted by lat/long
  fill(temperature_category, .direction = "down")

# Optionally, replace the remaining NAs with "Unknown"
merged_data <- merged_data %>%
  mutate(temperature_category = ifelse(is.na(temperature_category), "Unknown", temperature_category))

#re-order key
merged_data$temperature_category <- factor(merged_data$temperature_category, 
                                            levels = c("tropical", 
                                                       "temperate (warm)", 
                                                       "temperate (cool)"))
# #remove the duplicate TAS145 site
# merged_data <- merged_data[-1, ]

# Create the map
  map <- ggplot() +
  geom_sf(data = world, fill = "gray90", color = "gray50") +
  geom_sf(data = oz_states, fill = "gray90", color = "gray50") +
  geom_point(data = all_lats_longs_clean, 
             aes(x = longitude, y = latitude), 
             shape = 21, size = 1.8, fill = NA, color = "black") +  # Hollow points
  geom_point(data = coords_from_analysis_sites, 
             aes(x = longitude, y = latitude, fill = temperature_category), 
             size = 1.8, alpha = 0.9, shape = 21, color = "grey10") +  # Black borders
  scale_fill_manual(values = c("temperate (cool)" = "dodgerblue2", 
                               "temperate (warm)" = "darkgoldenrod1", 
                               "tropical" = "firebrick2")) +  # Fill colors
  xlab("Longitude") + 
  ylab("Latitude") +
  coord_sf(xlim = c(110.00, 160.00), ylim = c(-45.00, -8.00), expand = FALSE) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "bl", which_north = "true",
                         pad_x = unit(0.1, "in"), pad_y = unit(0.3, "in"),
                         style = north_arrow_fancy_orienteering) +
  labs(fill = "Biogeography") +  # Legend title for fill colors
  guides(fill = guide_legend(override.aes = list(size = 4))) +  # Larger legend points
  theme_minimal(base_size = 8) +
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text = element_text(size = 7),
        axis.title = element_text(size = 9))

# Print the map
print(map)

pdf("Final_map.pdf", width = 10, height = 7)  # Dimensions in inches
print(map)  # Replace 'map' with your plot object
dev.off()  # Close the device
```
