---
title: "Poorly studied reef species face elevated extinction risk"
author: "Olivia J. Johnson, Freddie J. Heather & Camille Mellin"
date: "`r Sys.Date()`"
output: 
  rmdformats::downcute:
  # bookdown::html_document2:
    code_folding: show
---

# About this script

The following document provides the R code for the analysis for the paper: Johnson *et al.* (in review.) Poorly studied reef species face elevated extinction risk.

**\*Corresponding author:**
[olivia.johsnon\@utas.edu.au](mailto:olivia.johsnon@utas.edu.au){.email}

+------------------------+-------------------------------------------------------------------------------------------------------+
| **R-version**          | 4.3.1 (2023-06-16 ucrt) -- "Beagle Scouts"                                                            |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **platform**           | x86_64-w64-mingw32                                                                                    |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article DOI**        |                                                                                                       |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article link**       |                                                                                                       |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Article citation**   | Johnson *et al.* (in review.) Poorly studied reef species face elevated extinction risk warming. |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Time series**        | 2008-2022                                                                                             |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Geographical scale** | Continental Australia                                                                                 |
+------------------------+-------------------------------------------------------------------------------------------------------+
| **Code author contact**| olivia.johnson@utas.edu.au                                                                            |
+------------------------+-------------------------------------------------------------------------------------------------------+

# Set-up

```{r knitr-options}
#| include = FALSE

library(knitr)
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

# save csv files for each step?
save_csv <- TRUE

```

# Loading packages
Load in all required packages for this analysis
```{r packages}

library(knitr)
library(tidyverse)
library(janitor)
library(lubridate)
library(lme4)
library(lmerTest)
library(zoo, include.only = "na.approx")
library(ggforce, include.only = "facet_col")

# "not in" function
`%!in%` <- Negate(`%in%`)
```

#Cleaning script
The code for cleaning the raw data is available in the R script "cleaning.R" - 
The raw data is in a wide format and is publicly available to download from 
https://portal.aodn.org.au/ - you will need to download this raw data before
proceeding with running the cleaning code. 

```{r}
#source("cleaning.R")
```

# Modelling and plots
The second half of this analysis runs a series of Linear Mixed Effects models to
understand population trends of reef species around continental Australia from
2008 - 2022, and then produce the plots as per those seen in the manuscript by
Johnson et al. (in review)

# Data import
Read in the cleanly produced data files that were created in the chunks above. 
```{r - read in all tidied csv files}
algae_data <- read_csv("algae_data.csv", show_col_types = FALSE)

fish_data <- read_csv("fish_data.csv", show_col_types = FALSE)

inverts_data <- read_csv("inverts_data.csv", show_col_types = FALSE)

coral_data <- read_csv("coral_data.csv", show_col_types = FALSE)
```

## Covariate data
Read in further data covariate data that will be required
```{r covariate-data-import}
#Species-level information
species_tbl <- 
  read_csv(file = "species_info.csv", 
           show_col_types = FALSE, 
           skip_empty_rows = TRUE) 

#Temperature time-series (post-2008) data for each site
temperature_raw <- 
  read_csv(file =  "temperature_timeseries.csv",
           show_col_types = FALSE, 
           skip_empty_rows = TRUE)

#Endemism dataframe
endemism <- 
  read_csv(file =  "endemism.csv",
           show_col_types = FALSE, 
           skip_empty_rows = TRUE)
```

# Linear mixed effects models
Below, I will perform a series of linear mixed-effects models on the
data above that has been cleaned. Each major group will be analysed
separately (algae, fish, inverts and coral) and a loop performed on the
model so that each individual species has their trends over time
analysed.

Filter parameters set above are the bare minimum that are required in
order to fit a linear line and therefore perform the linear
mixed-effects models on the data.

All models appear to run, but there are some warnings at the end of the
loop. Where issues arise in the modelling including the "isSingular"
error, indicating convergence when the model runs and within the data
for the species the model is being performed. The individual species
that are having this error are then inspected, and in some instances has
been randomly plotted to see (if there are any obvious signs) what may
be causing convergence. If it appears that it is limited data to
effectively perform the model, the model loop is re-run, with a flag to
remove any species that are having the "singular fit" error when the
loop is performed.

The table output is the list of species that has had their trends over
time analysed and the major result values needed to understand the
trends put into a table (slope, intercept, p-value, and significance of
the trend).

## Algae LMM
```{r}
result_algae <- tibble(
  species_name = character(),
  slope = numeric(),
  intercept = numeric(),
  p_value = numeric(),
  ci_low = numeric(),
  ci_high = numeric(),
  se = numeric(),
  sd_residuals = numeric(),
  aic = numeric(),
  bic = numeric(),
  error_flag = character()  # New column to track the error flag
) 


###add in check of minimum number of years and observations for individual species (add to supplementary material)
algae_data %>% 
  select(species_name, survey_year) %>% 
  distinct() %>% 
  mutate(range_yrs = max(survey_year)-min(survey_year), 
         .by = species_name) %>% 
  count(species_name, range_yrs)

for (i in unique(algae_data$species_name)) {
  
  mod_data_algae <- 
    algae_data %>% 
    filter(species_name == i)

# Check the number of unique levels of site_code for each species_name
  unique_levels <- n_distinct(algae_data$site_code)

  if (unique_levels <= 1) {
    warning(paste("Species", i, "has only one level of site_code. Consider using a different grouping variable."))
    next  # Skip to the next iteration
  }    
    
  tryCatch({
    
    algae_model <- 
    lmerTest::lmer(sqrt_cover ~ survey_year + (1|site_code), 
                   data = mod_data_algae) 
  
  intercept <- lme4::fixef(algae_model)[1] %>% as.numeric()
  slope <- lme4::fixef(algae_model)[2] %>% as.numeric()
  
  # Check if the model has a singular fit
  if (isSingular(algae_model)) {
    error_flag <- "Singular Fit"  # Set the error flag
  } else {
    error_flag <- ""  # No error flag
    
    # Extracting p-value
    p_value <- summary(algae_model)$coefficients[2, "Pr(>|t|)"]
    
    ci <- confint(algae_model, "survey_year")
    ci_low <- ci[1]
    ci_high <- ci[2]
    
    se <- summary(algae_model)$coefficients[2, "Std. Error"]
    sd_residuals <- summary(algae_model)$sigma
    aic <- AIC(algae_model)
    bic <- BIC(algae_model)
  }
  
  output <- summary(object = algae_model)
  
  result_algae <- 
    result_algae %>% 
    add_row(species_name = i, 
            slope = slope, 
            intercept = intercept, 
            p_value = p_value, 
            ci_low = ci_low, 
            ci_high = ci_high, 
            se = se, 
            sd_residuals = sd_residuals,
            aic = aic,
            bic = bic,
            error_flag = error_flag) %>%
    mutate(sig = case_when(p_value <= 0.001 ~ "***",
                           p_value <= 0.01  ~ "**",
                           p_value <= 0.05  ~ "*",
                           TRUE ~ ""))   }, error = function(e) {
    cat("Error occurred for species:", i, "\n")
    cat("Error message:", conditionMessage(e), "\n")
  })

  
  # print(summary(algae_model))
  
}

# Liv's previous comment:
#Model runs and produces results for 246 of the algae species - multiple species are having the error "boundary (singular) fit: see help('isSingular')" indicating there is a convergence issue with that species. Upon inspection of a few of the species where this error occurs in the model, there are only a very limited number of positive observations (>3 as set in the minimum parameters) in the data suggesting this could be causing issues in the algorithm's ability to find the optimal estimates of the model parameters. 

```

## Fish LMM
```{r}
result_fish <- tibble(
  species_name = character(),
  slope = numeric(),
  intercept = numeric(),
  p_value = numeric(),
  ci_low = numeric(),
  ci_high = numeric(),
  se = numeric(),
  sd_residuals = numeric(),
  aic = numeric(),
  bic = numeric(),
  error_flag = character()  # New column to track the error flag
) 


###add in check of minimum number of years and observations for individual species (add to supplementary material)
fish_data %>% 
  select(species_name, survey_year) %>% 
  distinct() %>% 
  mutate(range_yrs = max(survey_year)-min(survey_year), 
         .by = species_name) %>% 
  count(species_name, range_yrs) 


for (i in unique(fish_data$species_name)) {
  
  mod_data_fish <- 
    fish_data %>% filter(species_name == i)
#   
# # Check the number of unique levels of site_code for each species_name
#   unique_levels <- n_distinct(mod_data_fish$site_code) 
# 
# 
#   if (unique_levels <= 1) {
#     warning(paste("Species", i, "has only one level of site_code. Consider using a different grouping variable."))
#     next  # Skip to the next iteration
#   }  
  
  tryCatch({fish_model <- 
    lmerTest::lmer(log_count ~ year + (1|site_code), data = mod_data_fish) 
  
  intercept <- lme4::fixef(fish_model)[1] %>% as.numeric()
  slope <- lme4::fixef(fish_model)[2] %>% as.numeric()
  # print(i)
  
  # Check if the model has a singular fit
  if (isSingular(fish_model)) {
    error_flag <- "Singular Fit"  # Set the error flag
  } else {
    error_flag <- ""  # No error flag
    # Extracting p-value
    p_value <- summary(fish_model)$coefficients[2, "Pr(>|t|)"]
    
    ci <- confint(fish_model, "year")
    ci_low <- ci[1]
    ci_high <- ci[2]
    
    se <- summary(fish_model)$coefficients[2, "Std. Error"]
    sd_residuals <- summary(fish_model)$sigma
    aic <- AIC(fish_model)
    bic <- BIC(fish_model)
  }
  
  output <- summary(object = fish_model)
  
  result_fish <- result_fish %>% 
    add_row(species_name = i, 
            slope = slope, 
            intercept = intercept, 
            p_value = p_value, 
            ci_low = ci_low, 
            ci_high = ci_high, 
            se = se, 
            sd_residuals = sd_residuals,
            aic = aic,
            bic = bic,
            error_flag = error_flag) %>%
    mutate(sig = case_when(p_value <= 0.001 ~ "***",
                           p_value <= 0.01  ~ "**",
                           p_value <= 0.05  ~ "*",
                           TRUE ~ ""))  }, error = function(e) {
    cat("Error occurred for species:", i, "\n")
    cat("Error message:", conditionMessage(e), "\n")
  })
  
  #print(summary(fish_model))
  
}
```

## Mobile macro-invertebrates LMM
```{r}
result_inverts <- tibble(
  species_name = character(),
  slope = numeric(),
  intercept = numeric(),
  p_value = numeric(),
  ci_low = numeric(),
  ci_high = numeric(),
  se = numeric(),
  sd_residuals = numeric(),
  aic = numeric(),
  bic = numeric(),
  error_flag = character()  # New column to track the error flag
) 

###add in check of minimum number of years and observations for individual species (add to supplementary material)
inverts_data %>% 
  select(species_name, survey_year) %>% 
  distinct() %>% 
  mutate(range_yrs = max(survey_year)-min(survey_year), 
         .by = species_name) %>% 
  count(species_name, range_yrs) 

for (i in unique(inverts_data$species_name)) {
  
  mod_data_inverts <- 
    inverts_data %>% 
    filter(species_name == i)
  
# Check the number of unique levels of site_code for each species_name
  unique_levels <- n_distinct(inverts_data$site_code)  
  
    if (unique_levels <= 1) {
    warning(paste("Species", i, "has only one level of site_code. Consider using a different grouping variable."))
    next  # Skip to the next iteration
  }
  
  # Fit the mixed-effects model
  tryCatch({inverts_model <- lmerTest::lmer(log_count ~ survey_year + (1|site_code), data = mod_data_inverts) 
  
  intercept <- lme4::fixef(inverts_model)[1] %>% as.numeric()
  slope <- lme4::fixef(inverts_model)[2] %>% as.numeric()
  # print(i)
  
  # Check if the model has a singular fit
  if (isSingular(inverts_model)) {
    error_flag <- "Singular Fit"  # Set the error flag
  } else {
    error_flag <- ""  # No error flag
    # Extracting p-value
    p_value <- summary(inverts_model)$coefficients[2, "Pr(>|t|)"]
    
    ci <- confint(inverts_model, "survey_year")
    ci_low <- ci[1]
    ci_high <- ci[2]
    
    se <- summary(inverts_model)$coefficients[2, "Std. Error"]
    sd_residuals <- summary(inverts_model)$sigma
    aic <- AIC(inverts_model)
    bic <- BIC(inverts_model)
  }
  
  output <- summary(object = inverts_model)
  
  result_inverts <- result_inverts %>% 
    add_row(species_name = i, slope = slope, 
            intercept = intercept, 
            p_value = p_value,
            ci_low = ci_low, 
            ci_high = ci_high, 
            se = se, 
            sd_residuals = sd_residuals,
            aic = aic,
            bic = bic,
            error_flag = error_flag) %>%
    mutate(sig = case_when(p_value <= 0.001 ~ "***",
                           p_value <= 0.01  ~ "**",
                           p_value <= 0.05  ~ "*",
                           TRUE ~ "")) }, error = function(e) {
    cat("Error occurred for species:", i, "\n")
    cat("Error message:", conditionMessage(e), "\n")
  })
  
  # print(summary(inverts_model))
  
}
```

## Coral LMM
```{r}
result_coral <- tibble(
  species_name = character(),
  slope = numeric(),
  intercept = numeric(),
  p_value = numeric(),
  ci_low = numeric(),
  ci_high = numeric(),
  se = numeric(),
  sd_residuals = numeric(),
  aic = numeric(),
  bic = numeric(),
  error_flag = character()  # New column to track the error flag
) 

###add in check of minimum number of years and observations for individual species (add to supplementary material)
coral_data %>% 
  select(species_name, survey_year) %>% 
  distinct() %>% 
  mutate(range_yrs = max(survey_year)-min(survey_year), 
         .by = species_name) %>% 
  count(species_name, range_yrs) 


for (i in unique(coral_data$species_name)) {
  
  mod_data_coral <- coral_data %>% filter(species_name == i)
  
  
  # Check the number of unique levels of site_code for each species_name
  unique_levels <- n_distinct(coral_data$site_code)
  
  if (unique_levels <= 1) {
    warning(paste("Species", i, "has only one level of site_code. Consider using a different grouping variable."))
    next  # Skip to the next iteration
  }
  
  # Fit the mixed-effects model
  tryCatch({coral_model <- lmerTest::lmer(sqrt_pcover ~ survey_year + (1|site_code), data = mod_data_coral) 
  
  intercept <- lme4::fixef(coral_model)[1] %>% as.numeric()
  slope <- lme4::fixef(coral_model)[2] %>% as.numeric()
  # print(i)
  
  # Check if the model has a singular fit
  if (isSingular(coral_model)) {
    error_flag <- "Singular Fit"  # Set the error flag
  } else {
    error_flag <- ""  # No error flag
    # Extracting p-value
    p_value <- summary(coral_model)$coefficients[2, "Pr(>|t|)"]
    
    ci <- confint(coral_model, "survey_year")
    ci_low <- ci[1]
    ci_high <- ci[2]
    
    se <- summary(coral_model)$coefficients[2, "Std. Error"]
    sd_residuals <- summary(coral_model)$sigma
    aic <- AIC(coral_model)
    bic <- BIC(coral_model)
  }
  
  output <- summary(object = coral_model)
  
  result_coral <- result_coral %>% 
    add_row(species_name = i, slope = slope, 
            intercept = intercept, 
            p_value = p_value,
            ci_low = ci_low, 
            ci_high = ci_high, 
            se = se, 
            sd_residuals = sd_residuals,
            aic = aic,
            bic = bic,
            error_flag = error_flag) %>%
    mutate(sig = case_when(p_value <= 0.001 ~ "***",
                           p_value <= 0.01  ~ "**",
                           p_value <= 0.05  ~ "*",
                           TRUE ~ ""))  
  }, error = function(e) {
    cat("Error occurred for species:", i, "\n")
    cat("Error message:", conditionMessage(e), "\n")
  })
  
  # print(summary(coral_model))
  
}

```

# Rate of change
The slope and p-value do not mean much on their own - thus it is
important that a value is determined for the rate of change for each of
the species.

Below, the exact number of years a species has been measured that is
included in the model is determined. The slope is then divided by this
time portion, to give the rate of change of an individual species over 1
year (in both proportion and in a percentage form). This result is then
taken further (in the next chunk of code) and multiplied by 10 years -
as a measure of Criteria A "Population Size Reduction" of the IUCN Red
List Threatened Species Category (options are either 3 generations or 10
years - many of these species gave an unknown generation length). This
will then help determine the rate of change over a known number of years
for that species, and in turn, help to assign their conservation status
based on the IUCN Red List categories and criteria.
```{r rate-change}
#Algae
result_algae <- result_algae %>% 
  mutate(y_2008 = (intercept+slope*(2008))^2, 
         y_2021 = (intercept+slope*(2021))^2) %>% 
  mutate(diff = (y_2021-y_2008)) %>% 
  mutate(pc = (diff/y_2008)*100) %>% 
  mutate(decade_change = (pc/13)*10)

#Coral
result_coral <- result_coral %>% 
  mutate(y_2008 = (intercept+slope*(2008))^2, 
         y_2021 = (intercept+slope*(2021))^2) %>% 
  mutate(diff = (y_2021-y_2008)) %>% 
  mutate(pc = (diff/y_2008)*100) %>% 
  mutate(decade_change = (pc/13)*10)

#Fish
result_fish <- result_fish %>% 
  mutate(y_2008 = 10^(intercept+slope*(2008)), 
         y_2021 = 10^(intercept+slope*(2021))) %>% 
  mutate(diff = (y_2021-y_2008)) %>% 
  mutate(pc = (diff/y_2008)*100) %>% 
  mutate(decade_change = (pc/13)*10)

#Invertebrates
result_inverts <- result_inverts %>% 
  mutate(y_2008 = 10^(intercept+slope*(2008)), 
         y_2021 = 10^(intercept+slope*(2021))) %>% 
  mutate(diff = (y_2021-y_2008)) %>% 
  mutate(pc = (diff/y_2008)*100) %>% 
  mutate(decade_change = (pc/13)*10)
```

```{r - Chi Squared Test}
#ALL 638 species

# Create a matrix with your data
data <- matrix(c(136, 124, 197, 181), nrow = 2)

# Attach row and column names
rownames(data) <- c("Threatened", "Non-Threatened")
colnames(data) <- c("Endemic", "Non-Endemic")

# Print the data
print(data)

# Perform the chi-squared test
result <- chisq.test(data)

# Print the result
print(result)

# Extract and print the p-value
p_value <- result$p.value
print(p_value)

# Interpret the result
if (p_value < 0.05) {
  cat("There is a significant association between endemism status and threatened status.\n")
} else {
  cat("There is no significant association between endemism status and threatened status.\n")
}


##Significant declining species only 

# Create a matrix with your data
data <- matrix(c(47, 260, 31, 378), nrow = 2)

# Attach row and column names
rownames(data) <- c("Threatened", "Non-Threatened")
colnames(data) <- c("Endemic", "Non-Endemic")

# Print the data
print(data)

# Perform the chi-squared test
result <- chisq.test(data, correct = FALSE)

# Print the result
print(result)

# Extract and print the p-value
p_value <- result$p.value
print(p_value)

# Interpret the result
if (p_value < 0.05) {
  cat("There is a significant association between endemism status and threatened status.\n")
} else {
  cat("There is no significant association between endemism status and threatened status.\n")
}

#1 in 10 endemics are threatened - almost twice as greater proportion of endemics were threatened compared to non-endemics (0.056)
#27/297 and 20/408
```


#Plotting
Read in csv of LMM results, rate of change from above and the species current 
IUCN status.This spreadsheet has also filtered out all of the species that 
produced the 'isSingular' error in the modelling stage. 
```{r}
#Read in data
trending_species <- read_csv("LMM_output.csv", show_col_types = FALSE)

#Decade decline plots
trending_species$taxon <- factor(trending_species$taxon, levels = c("Macroalgae", "Invertebrate", "Vertebrate"))
trending_species$biogeog <- factor(trending_species$biogeog, levels = c("cool (temperate)", "warm (temperate)", "tropical"))

#Endemic species plot
trending_species %>% 
  mutate(taxon = case_when(taxon == "Vertebrate" ~ "c) Fishes",
                           taxon == "Invertebrate" ~ "b) Mobile macro-invertebrates",
                           taxon == "Macroalgae" ~ "a) Macroalgae")) %>%
  mutate(taxon = factor(taxon, levels = c("a) Macroalgae", "b) Mobile macro-invertebrates", "c) Fishes"))) %>%
  filter(p_value < 0.05, 
         slope < 0, 
         endemism == "endemic") %>% 
  ggplot(aes(x = decade_change, y = reorder(species_name, -decade_change))) +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -0, xmax = -30, fill = "#cccc33") +   
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -30, xmax = -50, fill = "yellow") +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -50, xmax = -80, fill = "orange") +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -80, xmax = -100, fill = "red") +
  geom_vline(xintercept = c(-30, -50, -80), lty = 2) +
  geom_errorbarh(aes(xmin = after_stat(x - stat(y)), xmax = after_stat(x + stat(y)), height = 0.2)) + 
  geom_point(aes(shape = biogeog), size = 4.5) + #try a bigger sized point
  #geom_point(size = 3) + #####
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    facet_wrap(~taxon, ncol = 1, scales = "free_y") +
  labs(
    x = "Rate of decline over a decade",
    y = "Species Name",
    shape = "") +
  theme_minimal() +
  theme(
    legend.position = "right",  
    panel.grid.major.y = element_blank(),  
    axis.title.y = element_blank(),  
    axis.text.y = element_text(face = "italic"), 
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 12, hjust = 0)
  )

#Non-endemic species plot
trending_species %>% 
  mutate(taxon = case_when(taxon == "Vertebrate" ~ "c) Fishes",
                           taxon == "Invertebrate" ~ "b) Mobile macro-invertebrates",
                           taxon == "Macroalgae" ~ "a) Macroalgae")) %>%
  mutate(taxon = factor(taxon, levels = c("a) Macroalgae", "b) Mobile macro-invertebrates", "c) Fishes"))) %>%
  filter(p_value < 0.05, 
         slope < 0, 
         endemism == "non-endemic") %>% 
  ggplot(aes(x = decade_change, y = reorder(species_name, -decade_change))) +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -0, xmax = -30, fill = "#cccc33") +   
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -30, xmax = -50, fill = "yellow") +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -50, xmax = -80, fill = "orange") +
  geom_rect(ymin = -Inf, ymax = Inf, xmin = -80, xmax = -100, fill = "red") +
  geom_vline(xintercept = c(-30, -50, -80), lty = 2) +
  geom_errorbarh(aes(xmin = after_stat(x - stat(y)), xmax = after_stat(x + stat(y)), height = 0.2)) + 
  geom_point(aes(shape = biogeog), size = 4.5) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
   ggforce::facet_col(taxon ~ ., scales = "free_y", space = "free") +
  labs(
    x = "Rate of decline over a decade",
    y = "Species Name",
    shape = "") +
  theme_minimal() +
  theme(
    legend.position = "right",  
    panel.grid.major.y = element_blank(),  
    axis.title.y = element_blank(),  
    axis.text.y = element_text(face = "italic"), 
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 11, hjust = 0, vjust = 1)
  )
```

#IUCN Status versus trend plots
```{r}
#Specify order of facet panels in plots
#desired_order <- c("Macroalgae", "Coral", "Invertebrate", "Vertebrate")

#ENDEMICS
all_species_endemic <- trending_species %>%
  filter(endemism == "endemic") %>%
  mutate(new = case_when(
    decade_change >= 0 ~ "Least Concern",
    decade_change >= -29.99 & decade_change < 0 ~ "Near Threatened",
    decade_change >= -49.99 & decade_change < -29 ~ "Vulnerable",
    decade_change >= -79.99 & decade_change < -50 ~ "Endangered",
    decade_change >= -100 & decade_change < -80 ~ "Critically Endangered"
  )) 

all_species_endemic  %>%
  count(taxon, IUCN, new, species_name) %>%
  mutate(new = factor(new, levels = c("Endangered", "Vulnerable", "Near Threatened", "Least Concern", "Lower Risk/conservation dependent"))) %>%
  mutate(taxon = case_when(taxon == "Vertebrate" ~ "d) Fishes",
                           taxon == "Invertebrate" ~ "c) Mobile macro-invertebrates",
                           taxon == "Macroalgae" ~ "a) Macroalgae",
                           taxon == "Coral" ~ "b) Corals")) %>%
  mutate(taxon = factor(taxon, levels = c("a) Macroalgae", "b) Corals", "c) Mobile macro-invertebrates", "d) Fishes"))) %>%
  ggplot(aes(x = IUCN, y = n, fill = new)) +
  geom_col() +
  facet_wrap(~taxon, ncol = 1,scales = "free_y") + 
  labs(x = "IUCN Red List Conservation Status",
       y = "Species Count",
       fill = "Population trend") +
  scale_fill_manual(
        values = c("Critically Endangered" = "red", "Endangered" = "orange", "Vulnerable" = "yellow", "Near Threatened" = "#cccc33", "Least Concern" = "green3"),
    breaks = c("Critically Endangered", "Endangered", "Vulnerable", "Near Threatened", "Least Concern")) +
  # scale_x_discrete(limits = c("Not Evaluated", "Data Deficient", "Near Threatened", "Least Concern", "Vulnerable", "Endangered")) +
  scale_x_discrete(limits = c("Not Evaluated", "Data Deficient", "Least Concern","Near Threatened",  "Vulnerable", "Endangered")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    legend.position = "right",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 11, hjust = 0),
    axis.title.y = element_text(margin = margin(r = 20)),
    strip.text.x = element_text(margin = margin(b = 10)),
    strip.text.y = element_text(margin = margin(r = 10)),
    axis.title.x = element_text(margin = margin(r = 40, t = 20))
  )

#NON-ENDEMICS
#Specify order of facet panels in plots
#desired_order <- c("Macroalgae", "Coral", "Invertebrate", "Vertebrate")

all_species_non_endemic <- trending_species %>%
  filter(endemism == "non-endemic") %>%
  mutate(new = case_when(
    decade_change >= 0 ~ "Least Concern",
    decade_change >= -29.99 & decade_change < 0 ~ "Near Threatened",
    decade_change >= -49.99 & decade_change < -29 ~ "Vulnerable",
    decade_change >= -79.99 & decade_change < -50 ~ "Endangered",
    decade_change >= -100 & decade_change < -80 ~ "Critically Endangered"
  )) 

all_species_non_endemic %>%
  count(taxon, IUCN, new, species_name) %>%
  mutate(new = factor(new, levels = c("Endangered", "Vulnerable", "Near Threatened", "Least Concern"))) %>%
  mutate(taxon = case_when(taxon == "Vertebrate" ~ "d) Fishes",
                           taxon == "Invertebrate" ~ "c) Mobile macro-invertebrates",
                           taxon == "Macroalgae" ~ "a) Macroalgae",
                           taxon == "Coral" ~ "b) Corals")) %>%
  mutate(taxon = factor(taxon, levels = c("a) Macroalgae", "b) Corals", "c) Mobile macro-invertebrates", "d) Fishes"))) %>%
  ggplot(aes(x = IUCN, y = n, fill = new)) +
  geom_col() +
  facet_wrap(~taxon, ncol = 1,scales = "free_y") + 
  labs(x = "IUCN Red List Conservation Status",
       y = "Species Count",
       fill = "Population trend") +
  scale_fill_manual(
    values = c("Critically Endangered" = "red", "Endangered" = "orange", "Vulnerable" = "yellow", "Near Threatened" = "#cccc33", "Least Concern" = "green3"),
    breaks = c("Critically Endangered", "Endangered", "Vulnerable", "Near Threatened", "Least Concern")) +
  # scale_x_discrete(limits = c("Not Evaluated", "Lower Risk/conservation dependent", "Data Deficient", "Least Concern", "Near Threatened", "Vulnerable", "Endangered")) +
  scale_x_discrete(limits = c("Not Evaluated",  "Data Deficient", "Lower Risk/conservation dependent","Least Concern", "Near Threatened", "Vulnerable", "Endangered")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    legend.position = "right",
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5),
    strip.text = element_text(size = 11, hjust = 0),
    axis.title.y = element_text(margin = margin(r = 20)),
    strip.text.x = element_text(margin = margin(b = 10)),
    strip.text.y = element_text(margin = margin(r = 10)),
    axis.title.x = element_text(margin = margin(r = 40, t = 20))
  )
```

#Produce map of sites where all data is available from around Australia
#MAP
```{r packages for mapping}
library("ggplot2")
library("sf")
library("rnaturalearth")
library("rnaturalearthdata")
library("ggspatial")
library("ozmaps")
library("rnaturalearthhires")
```

```{r - themes}
theme_set(theme_bw())
world <- ne_countries(scale = "medium", returnclass = "sf")
class(world)
```
#Determine latitude and longitude points
```{r - determine lats and longs - Clean}
#algae
algae_lats_longs_clean <- data.frame(Latitude = algae_data$latitude, Longitude = algae_data$longitude)

#fish
fish_lats_longs_clean <- data.frame(Latitude = fish_data$latitude, Longitude = fish_data$longitude)

#inverts
inverts_lats_longs_clean <- data.frame(Latitude = inverts_data$latitude, Longitude = inverts_data$longitude)

#coral
coral_lats_longs_clean <- data.frame(Latitude = coral_data$latitude, Longitude = coral_data$longitude)

# Combine latitude and longitude tables for algae, fish, inverts, and coral
all_lats_longs_clean <- bind_rows(algae_lats_longs_clean, fish_lats_longs_clean, inverts_lats_longs_clean, coral_lats_longs_clean)

```
#Produce map
```{r - lats and long code for map}
#RAW data map
coordinates <- all_lats_longs_clean

#how many sites in total are on this map?
unique_species_count <- length(unique(coordinates$Latitude))
unique_species_count


# Create a simple world map for context
world <- ne_countries(scale = "medium", returnclass = "sf")
oz_states <- ne_states(country = "Australia", returnclass = "sf")

# Plot the map with your data points
ggplot(data = world) +
  geom_sf(data = oz_states) +
  geom_point(data = coordinates, aes(x = Longitude, y = Latitude), color = "deepskyblue") +
  xlab("Longitude") + ylab("Latitude") +
  coord_sf(xlim = c(110.00, 160.00), ylim = c(-45.00, -8.00), expand = FALSE) +
  annotation_scale(location = "bl", width_hint = 0.3) +
  annotation_north_arrow(location = "bl", which_north = "true",
                         pad_x = unit(0.1, "in"), pad_y = unit(0.3, "in"),
                         style = north_arrow_fancy_orienteering) +
  theme_bw(base_size = 8)
```